Tagging

### 2007-05-08 slides/tag.tex
tweaks to confusion matrix code (replaced dict with set); initial notes on confusion matrix added to tag.txt
svn/trunk@4481

 g \pause
  \begin{itemize}
  \item Tell us what words are likely to occur in the neighbourhood
    (eg adjectives often followed by nouns, personal pronouns often
    followed by verbs, possessive pronouns by nouns)
  \item Pronunciations can be dependent on part of speech, eg
    \textcolor{green}{object, content, discount} (useful for speech
    synthesis and speech recognition)
  \item Can help information retrieval and extraction (stemming,
    partial parsing) 
  \item Useful component in many NLP systems
  \end{itemize}

@@

\begin{frame}
  \frametitle{Closed classes in English}
  \begin{tabular}{ll}
  \textbf{prepositions} & on, under, over, to, with, by\\
  \textbf{determiners} & the, a, an, some\\
  \textbf{pronouns} & she, you, I, who\\
  \textbf{conjunction}s & and, but, or, as, when, if\\
  \textbf{auxiliary} verbs & can, may, are\\
  \textbf{particles} & up, down, at, by\\
  \textbf{numeral}s & one, two, first, second\\
  \end{tabular}
\end{frame}

\begin{frame}
  \frametitle{Open classes}
  \begin{description}
  \item[nouns] Proper nouns (\textcolor{green}{Scotland}, \textcolor{green}{BBC}),\\
    common nouns:
    \begin{itemize}
    \item count nouns (\textcolor{green}{goat},
      \textcolor{green}{glass})
    \item mass nouns (\textcolor{green}{snow},
      \textcolor{green}{pacifism})
    \end{itemize}
  \item[verbs] actions and processes (\textcolor{green}{run},
    \textcolor{green}{hope}), also auxiliary verbs
  \item[adjectives] properties and qualities (age, colour, value)
  \item[adverbs] modify verbs, or verb phrases, or other adverbs:\\
    \textcolor{green}{\emph{Unfortunately} John walked \emph{home
        extremely slowly yesterday}}
  \end{description}
\end{frame}

@@

\begin{frame}
  \frametitle{The Penn Treebank tagset (1)}
  {\footnotesize
    \begin{tabular}{|lll|lll|}
      \hline
      CC & Coord Conjuncn & \emph{and,but,or} &  
      NN & Noun, sing. or mass & \emph{dog} \\
      CD & Cardinal number & \emph{one,two} & 
      NNS & Noun, plural & \emph{dogs} \\
      DT & Determiner & \emph{the,some} &  
      NNP & Proper noun, sing. & \emph{Edinburgh} \\
      EX & Existential there & \emph{there} &  
      NNPS & Proper noun, plural & \emph{Orkneys} \\
      FW & Foreign Word & \emph{mon dieu} &  
      PDT & Predeterminer & \emph{all, both}\\
      IN & Preposition & \emph{of,in,by} & 
      POS & Possessive ending & \emph{'s} \\
      JJ & Adjective & \emph{big} & 
      PP & Personal pronoun & \emph{I,you,she} \\
      JJR & Adj., comparative & \emph{bigger} & 
      PP\$ & Possessive pronoun & \emph{my,one's} \\
      JJS & Adj., superlative & \emph{biggest} &  
      RB & Adverb & \emph{quickly} \\
      LS & List item marker & \emph{1,One} &  
      RBR & Adverb, comparative & \emph{faster} \\
      MD & Modal & \emph{can,should} &
      RBS & Adverb, superlative & \emph{fastest} \\
      \hline
    \end{tabular}
  }
\end{frame}

\begin{frame}
  \frametitle{The Penn Treebank tagset (2)}
  {\footnotesize
    \begin{tabular}{|lll|lll|}
      \hline
      RP & Particle & \emph{up,off} &
      WP\$ & Possessive-Wh & \emph{whose} \\ 
      SYM & Symbol & \emph{+,\%,\&}  &
      WRB & Wh-adverb & \emph{how,where} \\
      TO & ``to'' & \emph{to} &
      \$ & Dollar sign & \emph{\$}\\
      UH & Interjection & \emph{oh, oops} &
      \# & Pound sign & \emph{\#} \\
      VB & verb, base form &     \emph{eat} &
      `` & Left quote & ` , `` \\
      VBD & verb, past tense & \emph{ate} &
      '' & Right quote & ', '' \\
      VBG & verb, gerund & \emph{eating} &
      ( & Left paren & ( \\
      VBN & verb, past part & \emph{eaten} &
      ) & Right paren & ) \\
      VBP & Verb, non-3sg, pres & \emph{eat} &
      , & Comma & , \\
      VBZ & Verb, 3sg, pres & \emph{eats} &
      . & Sent-final punct & . ! ? \\
      WDT & Wh-determiner &  \emph{which,that} &
      : & Mid-sent punct. & : ; --- ... \\
      WP & Wh-pronoun & \emph{what,who} & & & \\
      \hline
    \end{tabular}
  }
\end{frame}

@@

\begin{frame}
  \frametitle{Information sources for tagging}

  What information can help decide the correct PoS tag for a word?
  \begin{description}
  \item[Other PoS tags] Even though the PoS tags of other words may be
    uncertain too, we can use information that some tag sequences are
    more likely than others (eg \emph{the/AT red/JJ drink/NN} vs \emph{the/AT
      red/JJ drink/VBP}).\newline
    Using \emph{only} information about the most likely PoS tag
    sequence does not result in an accurate tagger (about 77\%
    correct)
  \item[The word identity] Many words can gave multiple possible tags,
    but some are more likely than others (eg \emph{fall/VBP} vs
    \emph{fall/NN}) \newline
    Tagging each word with its most common tag results in a tagger
    with about 90\% accuracy
  \end{description}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Tagging in NLTK}
The simplest possible tagger tags everything as a noun:
{\small
\begin{verbatim}
from nltk_lite import tokenize
text = 'There are 11 players in a football team'
text_tokens = list(tokenize.whitespace(text))
# ['There', 'are', '11', 'players', 'in', 'a', 'football', 'team']
\end{verbatim}
  \pause
\begin{verbatim}
from nltk_lite import tag
mytagger = tag.Default('nn')
for t in mytagger.tag(text_tokens):
    print t
# ('There', 'NN')
# ('are', 'NN')
# ...
\end{verbatim}}
\end{frame}

\begin{frame}[fragile]
  \frametitle{A regular expression tagger}
We can use regular expressions to tag tokens based on regularities in
the text, eg numerals:

{\small
\begin{verbatim}
default_pattern = (r'.*', 'NN')
cd_pattern = (r' ^[0-9]+(.[0-9]+)?$', 'CD')
patterns = [cd_pattern, default_pattern]
NN_CD_tagger = tag.Regexp(patterns)
re_tagged = list(NN_CD_tagger.tag(text_tokens))
# [('There', 'NN'), ('are', 'NN'), ('11', 'NN'), ('players', 'NN'), 
('in', 'NN'), ('a', 'NN'), ('football', 'NN'), ('team', 'NN')]
\end{verbatim}
}%$
\end{frame}

>>>>

  \pause
  \begin{itemize}
  \item Tell us what words are likely to occur in the neighbourhood
    (eg adjectives often followed by nouns, personal pronouns often
    followed by verbs, possessive pronouns by nouns)
  \item Pronunciations can be dependent on part of speech, eg
    \textcolor{green}{object, content, discount} (useful for speech
    synthesis and speech recognition)
  \item Can help information retrieval and extraction (stemming,
    partial parsing) 
  \item Useful component in many NLP systems
  \end{itemize}

@@

\begin{frame}
  \frametitle{Closed classes in English}
  \textcolor{black}{
  \begin{tabular}{ll}
  \textbf{prepositions} & on, under, over, to, with, by\\
  \textbf{determiners} & the, a, an, some\\
  \textbf{pronouns} & she, you, I, who\\
  \textbf{conjunction}s & and, but, or, as, when, if\\
  \textbf{auxiliary verbs} & can, may, are\\
  \textbf{particles} & up, down, at, by\\
  \textbf{numeral}s & one, two, first, second\\
  \end{tabular}}
\end{frame}

\begin{frame}
  \frametitle{Open classes}
  \textcolor{black}{
  \begin{description}
  \item[nouns] Proper nouns (\textcolor{green}{Scotland}, \textcolor{green}{BBC}),\\
    common nouns:
    \begin{itemize}
    \item count nouns (\textcolor{green}{goat},
      \textcolor{green}{glass})
    \item mass nouns (\textcolor{green}{snow},
      \textcolor{green}{pacifism})
    \end{itemize}
  \item[verbs] actions and processes (\textcolor{green}{run},
    \textcolor{green}{hope}), also auxiliary verbs
  \item[adjectives] properties and qualities (age, colour, value)
  \item[adverbs] modify verbs, or verb phrases, or other adverbs:\\
    \textcolor{green}{\emph{Unfortunately} John walked \emph{home
        extremely slowly yesterday}}
  \end{description}}
\end{frame}

@@

\begin{frame}
  \frametitle{The Penn Treebank tagset (1)}
  \textcolor{black}{\footnotesize
    \begin{tabular}{|lll|lll|}
      \hline
      CC & Coord Conjuncn & \emph{and,but,or} &  
      NN & Noun, sing. or mass & \emph{dog} \\
      CD & Cardinal number & \emph{one,two} & 
      NNS & Noun, plural & \emph{dogs} \\
      DT & Determiner & \emph{the,some} &  
      NNP & Proper noun, sing. & \emph{Edinburgh} \\
      EX & Existential there & \emph{there} &  
      NNPS & Proper noun, plural & \emph{Orkneys} \\
      FW & Foreign Word & \emph{mon dieu} &  
      PDT & Predeterminer & \emph{all, both}\\
      IN & Preposition & \emph{of,in,by} & 
      POS & Possessive ending & \emph{'s} \\
      JJ & Adjective & \emph{big} & 
      PP & Personal pronoun & \emph{I,you,she} \\
      JJR & Adj., comparative & \emph{bigger} & 
      PP\$ & Possessive pronoun & \emph{my,one's} \\
      JJS & Adj., superlative & \emph{biggest} &  
      RB & Adverb & \emph{quickly} \\
      LS & List item marker & \emph{1,One} &  
      RBR & Adverb, comparative & \emph{faster} \\
      MD & Modal & \emph{can,should} &
      RBS & Adverb, superlative & \emph{fastest} \\
      \hline
    \end{tabular}
  }
\end{frame}

\begin{frame}
  \frametitle{The Penn Treebank tagset (2)}
  \textcolor{black}{\footnotesize
    \begin{tabular}{|lll|lll|}
      \hline
      RP & Particle & \emph{up,off} &
      WP\$ & Possessive-Wh & \emph{whose} \\ 
      SYM & Symbol & \emph{+,\%,\&}  &
      WRB & Wh-adverb & \emph{how,where} \\
      TO & ``to'' & \emph{to} &
      \$ & Dollar sign & \emph{\$}\\
      UH & Interjection & \emph{oh, oops} &
      \# & Pound sign & \emph{\#} \\
      VB & verb, base form &     \emph{eat} &
      `` & Left quote & ` , `` \\
      VBD & verb, past tense & \emph{ate} &
      '' & Right quote & ', '' \\
      VBG & verb, gerund & \emph{eating} &
      ( & Left paren & ( \\
      VBN & verb, past part & \emph{eaten} &
      ) & Right paren & ) \\
      VBP & Verb, non-3sg, pres & \emph{eat} &
      , & Comma & , \\
      VBZ & Verb, 3sg, pres & \emph{eats} &
      . & Sent-final punct & . ! ? \\
      WDT & Wh-determiner &  \emph{which,that} &
      : & Mid-sent punct. & : ; --- ... \\
      WP & Wh-pronoun & \emph{what,who} & & & \\
      \hline
    \end{tabular}
  }
\end{frame}

@@

\begin{frame}
  \frametitle{Information sources for tagging}
  \textcolor{black}{
  What information can help decide the correct PoS tag for a word?
  \begin{description}
  \item[Other PoS tags] Even though the PoS tags of other words may be
    uncertain too, we can use information that some tag sequences are
    more likely than others (eg \emph{the/AT red/JJ drink/NN} vs \emph{the/AT
      red/JJ drink/VBP}).\newline
    Using \emph{only} information about the most likely PoS tag
    sequence does not result in an accurate tagger (about 77\%
    correct)
  \item[The word identity] Many words can gave multiple possible tags,
    but some are more likely than others (eg \emph{fall/VBP} vs
    \emph{fall/NN}) \newline
    Tagging each word with its most common tag results in a tagger
    with about 90\% accuracy
  \end{description}}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Tagging in NLTK}
\textcolor{black}{The simplest possible tagger tags everything as a noun:}
\small
\begin{verbatim}
from nltk_lite import tokenize
text = 'There are 11 players in a football team'
text_tokens = list(tokenize.whitespace(text))
# ['There', 'are', '11', 'players', 'in', 'a', 'football', 'team']
\end{verbatim}
  \pause
\begin{verbatim}
from nltk_lite import tag
mytagger = tag.Default('nn')
for t in mytagger.tag(text_tokens):
    print t
# ('There', 'NN')
# ('are', 'NN')
# ...
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
  \frametitle{A regular expression tagger}
\textcolor{black}{
We can use regular expressions to tag tokens based on regularities in
the text, eg numerals:}
\small
\begin{verbatim}
default_pattern = (r'.*', 'NN')
cd_pattern = (r' ^[0-9]+(.[0-9]+)?$', 'CD')
patterns = [cd_pattern, default_pattern]
NN_CD_tagger = tag.Regexp(patterns)
re_tagged = list(NN_CD_tagger.tag(text_tokens))
# [('There', 'NN'), ('are', 'NN'), ('11', 'NN'), ('players', 'NN'), 
('in', 'NN'), ('a', 'NN'), ('football', 'NN'), ('team', 'NN')]
\end{verbatim}
\end{frame}
